---
title: "MA 710 - Assignment 2"
author: "MA 710 - Dataminers"
date: "February 22, 2017"
output:
  pdf_document: default
  html_document: default
---

### Loading libraries and data
```{r, message=FALSE, warning=FALSE}
library(dplyr)
library(readr)
library(arules)
```

We read in the csv file `MERGED2014_15_PP.csv` into the data frame `csb.df` using the `read_csv` function of the `readr` package.

```{r,message=FALSE, warning=FALSE, cache=TRUE}

year = 2000
csb.df = data.frame()

for (i in c(1:15)) {
  if (i<10){
    num = paste0("0", i, sep="")
  }
    else {
      num = i
    }
  file_name = paste0('MERGED',year,"_", num, "_PP.csv", sep="")
  csb.temp.df = read_csv(paste0('C:/Users/sevda/Documents/Bentley/710/Assignment 1/CollegeScorecard_Raw_Data/CollegeScorecard_Raw_Data/', file_name))
  csb.temp.df$YEAR = year
  csb.df = rbind(csb.df, csb.temp.df)
  year = year+1
  
}

```

### Data cleaning and variable modification

Based on our decision to select certain variables, we will use the `SELECT` command from `dply` package to create a smaller data frame which contains only the variables of interest.
```{r, message=FALSE, warning=FALSE}
csb.df %>%
  select(CONTROL, ST_FIPS, FAMINC, FEMALE, MARRIED,DEP_STAT_PCT_IND, FIRST_GEN, PREDDEG, CDR3, CDR3_DENOM,PCTPELL, PCTFLOAN, 
         DEBT_N, CUML_DEBT_N, HIGHDEG, REGION, NUMBRANCH, CCBASIC, CURROPER, YEAR) %>%
  {.} -> csb.vars.df
```


To prepare data for association rules, we need to create factor varibles from all of the numeric variables we chose to work with. Because firstly we need to conver the variable type into `factor` and then use the `make.ntiles` function, here we defined a new function `numeric_to_factor` to combine two steps together.
```{r, message=FALSE, warning=FALSE}
make.ntiles = function (inputvar, n) {
  inputvar %>%
    quantile(.,
             (1/n) * 1:(n-1),
             na.rm=TRUE
             ) %>%
    c(-Inf, ., Inf) %>%
    cut(inputvar,
        breaks=.,
        paste("Q", 1:n, sep="")
        )
}

numeric_to_factor <- function(var){
  var.temp= as.numeric(var)
  var.F=make.ntiles(var.temp, 3)
  return(var.F)
}

csb.vars.df %>%
  mutate(FAMINC=numeric_to_factor(FAMINC),
         FEMALE=numeric_to_factor(FEMALE),
         MARRIED=numeric_to_factor(MARRIED),
         DEP_STAT_PCT_IND=numeric_to_factor(DEP_STAT_PCT_IND),
         FIRST_GEN=numeric_to_factor(FIRST_GEN),
         CDR3=numeric_to_factor(CDR3),
         CDR3_DENOM=numeric_to_factor(CDR3_DENOM),
         PCTPELL=numeric_to_factor(PCTPELL),
         PCTFLOAN=numeric_to_factor(PCTFLOAN),
         DEBT_N=numeric_to_factor(DEBT_N),
         CUML_DEBT_N=numeric_to_factor(CUML_DEBT_N),
         NUMBRANCH=as.factor(NUMBRANCH),
         YEAR = as.factor(YEAR),
         CURROPER = as.factor(CURROPER),
         CCBASIC = as.factor(CCBASIC),
         HIGHDEG = as.factor(HIGHDEG)
        ) %>%
  {.} -> csb.fact.df
```
For the last step, we need to make sure that all variables we used here are categorical variables using following code:
```{r}
csb.fact.df %>%
  sapply(as.factor) %>%
  as.data.frame %>%
  {.} -> csb.var.df
```

The data frame `csb.var.df` contains our final set of variables, ready for further analysis.

## Association rules for the `CONTROL` target variable

The next step is to create the association rules. We will use the apriori function from the arules package. The target variable is CONTROL (ownership) which is a factor variable with three levels: 1 - Public college, 2 - Private college not for profit, 3 - Private college for profit. We will explroe rules which contain CONTROL on the right hand side of the rule. Our focus will be on maximum length of 10 and miniumum length of 2 and on confidence level of at least 0.07.

```{r, warning=FALSE}
csb.var.df %>%
   select(-PREDDEG,-FEMALE) %>%
  {.} -> csb.var.control.df

apriori.appearance = list(rhs=c('CONTROL=1','CONTROL=2', "CONTROL=3"), default='lhs')
apriori.parameter = list(support=0.01, confidence=0.07, minlen=2, maxlen=10)
apriori.control = list(verbose=FALSE)

rules.control = apriori(csb.var.control.df, parameter=apriori.parameter, appearance=apriori.appearance, control=apriori.control)


#rules = apriori(csb.var.control.df, parameter = list(supp = 0.1, conf = 0.7, minlen = 2, maxlen = 10))
#print(paste0("Number of rules:", length(rule.control)))
```
Based on our criteria of looking at only rules with confidence level higher that 0.07, support higher than 0.01, length between 2 and 10 and having the CONTROL variable on the right hand side, we come up wih 18,365 rules. We will sort the rules based on three separate criterie - support, confidence and lift.

### Association rules sorted by lift

First, we will do the sort by lift.

```{r}
inspect(sort(rules.control, by='lift')[1:3])

```

From the first rule, we can see that 1% of the cases are where the family income is in the highest quantile, the share of independent students is in the lowest quantile, the number of students in the mean debt cohost is in the medium quantile, the highest degree awarded is Bachelor, the region of the college is Great Lakes, the number of branches is 1 and the college is private not for profit. 96.98% of the colleges for which the the family income is in the highest quantile, the share of independent students is in the lowest quantile, the number of students in the mean debt cohost is in the medium quantile, the highest degree awarded is Bachelor, the region of the college is Great Lakes and the number of branches is 1 are private not for profit colleges. The number of private non profit colleges that have all of the above conditions (the family income is in the highest quantile, the share of independent students is in the lowest quantile, the number of students in the mean debt cohost is in the medium quantile, the highest degree awarded is Bachelor, the region of the college is Great Lakes, the number of branches is 1) is 252% higher than we would expect if the college is non for profit and the rest of the conditions were independent.

From the second rule, we can see that 1% of the total cases are where the family income is in the highest quantile, the share of independent students is in the lowest quantile, the number of students in the cummulative debt cohost is in the medium quantile, the highest degree awarded is Graduate, the region of the college is Great Lakes, the number of branches is 1 and the college is private not for profit. 96.98% of the colleges for which the the family income is in the highest quantile, the share of independent students is in the lowest quantile, the number of students in the cummulative debt cohost is in the medium quantile, the highest degree awarded is Graduate, the region of the college is Great Lakes and the number of branches is 1 are private not for profit colleges. The number of private non profit colleges that have all of the above conditions (the family income is in the highest quantile, the share of independent students is in the lowest quantile, the number of students in the cummulative debt cohost is in the medium quantile, the highest degree awarded is Graduate, the region of the college is Great Lakes, the number of branches is 1) is 252% higher than we would expect if the college is non for profit and the rest of the conditions were independent.

### Association rules sorted by support

The next step is to sort by the support values:

```{r}
inspect(sort(rules.control, by='support')[1:3])

```

From the first rule, we can see that 39% of the cases are where the Carnegie Basic classification is missing and the college is private for profit. 42.31% of the colleges for which the Carnegie basic classification is missing, are private for profit colleges. The number of private for profit colleges that have the left hand condition (the Carnegie Basic Classification is missing) is 1% lower than we would expect if the college is non for profit and the rest of the conditions were independent.

From the second rule, we can see that 39% of the cases are where the Current Operation information is missing and the college is private for profit. 42.31% of the colleges for which the Current Operation information is missing, are private for profit colleges. The number of private for profit colleges that have the left hand condition (the current operation information is missing) is 1% lower than we would expect if the college is non for profit and the rest of the conditions were independent.

From the third rule, we can see that 39% of the cases are where the Carnegie Basic classification and the Current Operation information is missing and the college is private and for profit. 42.31% of the colleges for which the Carnegie Basic classification and the Current Operation information are both missing, are private for profit colleges. The number of private non profit colleges that have the left hand condition (Carnegie Basic classification and the Current Operation information) is 1% lower than we would expect if the college is non for profit and the rest of the conditions were independent.


### Association rules sorted by confidence

The final step is to sort by the confidence value:
```{r}
inspect(sort(rules.control, by='confidence')[1:3])

```

From the first rule, we can see that 1.11% of the total cases are where the family income is in the middle quantile, percent of students who receive federal grants is in the lowest quantile, the number of students in the cummulative debt cohost is in the highest quantile, the highest degree awarded is Associate, the number of branches is 1 and the college is public. 99.74% of the colleges for which the the family income is in the middle quantile, percent of students who receive federal grants is in the lowest quantile, the number of students in the cummulative debt cohost is in the highest quantile, the highest degree awarded is Associate, the number of branches is 1 are public colleges. The number of public colleges that have all of the above left hand conditions (the family income is in the middle quantile, percent of students who receive federal grants is in the lowest quantile, the number of students in the cummulative debt cohost is in the highest quantile, the highest degree awarded is Associate, the number of branches is 1) is 235.53% higher than we would expect if the college is public and the rest of the conditions were independent.

From the first rule, we can see that 1.11% of the total cases are where the family income is in the middle quantile, percent of students who receive federal grants is in the lowest quantile, the number of students in the median debt cohost is in the highest quantile, the highest degree awarded is Associate, the number of branches is 1 and the college is public. 99.74% of the colleges for which the the family income is in the middle quantile, percent of students who receive federal grants is in the lowest quantile, the number of students in the median debt cohost is in the highest quantile, the highest degree awarded is Associate, the number of branches is 1 are public colleges. The number of public colleges that have all of the above left hand conditions (the family income is in the middle quantile, percent of students who receive federal grants is in the lowest quantile, the number of students in the median debt cohost is in the highest quantile, the highest degree awarded is Associate, the number of branches is 1) is 235.53% higher than we would expect if the college is public and the rest of the conditions were independent.


From the first rule, we can see that 1.11% of the total cases are where the family income is in the middle quantile, percent of students who receive federal grants is in the lowest quantile, the number of students in the cummulative debt cohost is in the highest quantile, the number of students in the median debt cohost is in the highest quantile, the highest degree awarded is Associate, the number of branches is 1 and the college is public. 99.74% of the colleges for which the the family income is in the middle quantile, percent of students who receive federal grants is in the lowest quantile, the number of students in the cummulative debt cohost is in the highest quantile, the number of students in the median debt cohost is in the highest quantile, the highest degree awarded is Associate, the number of branches is 1 are public colleges. The number of public colleges that have all of the above left hand conditions (the family income is in the middle quantile, percent of students who receive federal grants is in the lowest quantile, the number of students in the cummulative debt cohost is in the highest quantile, the number of students in the median debt cohost is in the highest quantile, the highest degree awarded is Associate, the number of branches is 1) is 235.53% higher than we would expect if the college is public and the rest of the conditions were independent.


## Association rules for the `PREDDEG` target variable

Drop the other two target variables `CONTROL` and `FEMALE`

```{r}
csb.var.df %>%
  select(-CONTROL,-FEMALE) %>%
  {.} -> csb.deg.df
```

The next step is to create the association rules for the `PREDDEG` variable. We will use the apriori function from the arules package. The parameters of the function are set to define the minimum support, minimum confidence, minimum length and maximum length for the association rules.

```{r}
apriori.appearance = list(rhs=c('PREDDEG=0','PREDDEG=1','PREDDEG=2','PREDDEG=3','PREDDEG=4' ),default='lhs')
apriori.parameter = list(support=0.01,confidence=0.07, minlen=2, maxlen=10)
apriori.control = list(verbose=FALSE)
rules.deg = apriori(csb.deg.df, parameter=apriori.parameter, appearance=apriori.appearance, control=apriori.control)
```

Let's first examine the number of rules we found for the parameters specified above.

```{r}
length(rules.deg)
```

The output tells that the function found 6863 association rules for the variable `PREDDEG`.

### Association rules sorted by support

```{r}
rules.by.support.deg = sort(rules.deg, by='support')
```

```{r}
inspect(rules.by.support.deg[1:3])
```

Rule 1: The first rule tells us that 39% of all institutions are predominantly Certificate-degree institutions with missing values for the variable `CURROPER`.

Rule 2: The second rule tells us that 39% of all institutions are predominantly Certificate-degree institutions with missing values for the variable `CCBASIC`.

Rule 3: The third rule tells us that 39% of all institutions are predominantly Certificate-degree institutions with missing values for the variables `CURROPER` and `CCBASIC`.

### Association rules sorted by confidence

```{r}
rules.by.conf.deg = sort(rules.deg, by='confidence')
``` 

```{r}
inspect(rules.by.conf.deg[1:3])
```

Rule 1: The first rule tells us that all those institutions in Puerto Rico (`ST_FIPS` = 72) in which the highest degree awarded is a Certificate degree are also predominantly certificate-degree granting institutions.

Rule 2: The second rule tells us that all those institutions in Connecticut (`ST_FIPS` = 9) in which the highest degree awarded is a Certificate degree are also predominantly certificate-degree granting institutions.

Rule 3: The third rule tells us that all those institutions in New Jersey (`ST_FIPS` = 34) in which the highest degree awarded is a Certificate degree are also predominantly certificate-degree granting institutions.

### Association rules sorted by lift

```{r}
rules.by.lift.deg = sort(rules.deg, by='lift')
```

```{r}
inspect(rules.by.lift.deg[1:3])
```

Rule 1: The first rule tells us that the number of unclassified institutions that have the highest number of students in the cumulative debt cohort and are non-degree granting is 23% higher that we would expect if the they were independent.

Rule 2: The second rule tells us that the number of unclassified institutions that have the highest number of students in the median debt cohort and are non-degree granting is 23% higher that we would expect if the they were independent.

Rule 3: The third rule tells us that the number of unclassified institutions that have the highest number of students in the cumulative debt cohort, highest number of students in the median debt cohort and are non-degree granting is 23% higher that we would expect if the they were independent.

## Association rules for the `FEMALE` target variable



The first code describes the association rules between FEMALE=Q1 and other variables. 
```{r}
csb.var.df %>%
 select(FEMALE, ST_FIPS, FAMINC, MARRIED,DEP_STAT_PCT_IND, FIRST_GEN, CDR3, CDR3_DENOM,PCTPELL, PCTFLOAN,
        DEBT_N, CUML_DEBT_N, HIGHDEG, REGION, NUMBRANCH, CCBASIC, CURROPER, YEAR) %>%
 {.} -> csb.var.female.df

apriori.appearance = list(rhs=c("FEMALE=Q1","FEMALE=Q2","FEMALE=Q3"), default='lhs')
apriori.parameter = list(support=0.01, confidence=0.07, minlen=2, maxlen=10)
apriori.control= list(verbose=FALSE)

rules.female = apriori(csb.var.female.df,
                       parameter=apriori.parameter,
                       appearance=apriori.appearance,
                       control = apriori.control)

inspect(sort(rules.female, by='lift')[1:3])
```

### Association rules sorted by confidence

The output indicates that the presence of NUMBRANCH=142, or MARRIED=Q8 and NUMBRANCH=142, or DEBT_N=Q10 and NUMBRANCH=142 makes it more likely that FEMALE=Q1; NUMBRANCH=142, MARRIED=Q8 and DEBT_N=Q10 are highly related to FEMALE=Q1. 
```{r}
csb.var.df %>%
 select(FEMALE, ST_FIPS, FAMINC, MARRIED,DEP_STAT_PCT_IND, FIRST_GEN, CDR3, CDR3_DENOM,PCTPELL, PCTFLOAN,
        DEBT_N, CUML_DEBT_N, HIGHDEG, REGION, NUMBRANCH, CCBASIC, CURROPER, YEAR) %>%
 {.} -> csb.var.female.df

apriori.appearance = list(rhs=c("FEMALE=Q1","FEMALE=Q2","FEMALE=Q3"), default='lhs')
apriori.parameter = list(support=0.01, confidence=0.07, minlen=2, maxlen=10)
apriori.control= list(verbose=FALSE)

rules.female = apriori(csb.var.female.df,
                       parameter=apriori.parameter,
                       appearance=apriori.appearance,
                       control = apriori.control)

inspect(sort(rules.female, by='confidence')[1:3])
```

The output rules are sorted by confidence. This result indicates that the probability of NUMBRANCH=142 contains FEMALE=Q1 is 1. The probability of MARRIED=Q8 and NUMBRANCH=142 contains FEMALE=Q1 is 1 and the probability of DEBT_N=Q10 and NUMBRANCH=142 contains FEMALE=Q1 is 1. From the output, we can infer that when NUMBRANCH=142, the FEMALE=Q1 will happen. 


### Association rules sorted by support

The code below shows the association rules sorted by support. 
```{r}
csb.var.df %>%
 select(FEMALE, ST_FIPS, FAMINC, MARRIED,DEP_STAT_PCT_IND, FIRST_GEN, CDR3, CDR3_DENOM,PCTPELL, PCTFLOAN,
        DEBT_N, CUML_DEBT_N, HIGHDEG, REGION, NUMBRANCH, CCBASIC, CURROPER, YEAR) %>%
 {.} -> csb.var.female.df

apriori.appearance = list(rhs=c("FEMALE=Q1","FEMALE=Q2","FEMALE=Q3"), default='lhs')
apriori.parameter = list(support=0.01, confidence=0.07, minlen=2, maxlen=10)
apriori.control= list(verbose=FALSE)

rules.female = apriori(csb.var.female.df,
                       parameter=apriori.parameter,
                       appearance=apriori.appearance,
                       control = apriori.control)

inspect(sort(rules.female, by='support')[1:3])
```
The output indicates that the probability of YEAR=2014 and FEMALE=Q1 is 0.0826; the probability of YEAR=2014 and FEMALE=Q2 is 0.0824 and the probability of YEAR=2014 and FEMALE=Q3 is 0.0824. The association rules are not strong enough between FEMALE and YEAR as well as FEMALE and other variables. 

### Association rules sorted by lift

Since lift is more than 1 in many association rules, the output are sorted by lift to see how FEMALE=Q2 and FEMALE=Q3 related to other variables. 
```{r}
apriori.appearance = list(rhs=c("FEMALE=Q2"), default='lhs')
apriori.parameter = list(support=0.01, confidence=0.07, minlen=2, maxlen=10)
apriori.control= list(verbose=FALSE)

rules.female = apriori(csb.var.female.df,
                       parameter=apriori.parameter,
                       appearance=apriori.appearance,
                       control = apriori.control)

inspect(sort(rules.female, by='lift')[1:3])
```
The lift results are all more than one. This indicates that CCBASIC=22, CURROPER=1 and YEAR=2014 are highly associated with FEMALE=2. When CCBASIC=22 and CURROPER=1, it is more likely that FEMALE=Q2; when CCBASIC=22, CURROPER=1 and YEAR=2-14, it is more likely that FEMALE=Q2. CCBASIC=22 makes it more likely that FEMALE=Q2. 


The third code describes the association rules between FEMALE=Q2 and other variabels.
```{r}
apriori.appearance = list(rhs=c("FEMALE=Q3"), default='lhs')
apriori.parameter = list(support=0.01, confidence=0.07, minlen=2, maxlen=10)
apriori.control= list(verbose=FALSE)

rules.female = apriori(csb.var.female.df,
                       parameter=apriori.parameter,
                       appearance=apriori.appearance,
                       control = apriori.control)

inspect(sort(rules.female, by='confidence')[1:3])
```
The results of lift are all more than one, which means that FAMINC=Q10, DEP_STAT_PCT_IND=Q1, FIRST_GEN=Q1, CURROPER=1 and YEAR=2014 are highly related to FEMALE=Q3. Among all the rules. FAMINC=Q10, DEP_STAT_PCT_IND=Q1 and FIRST_GEN=Q1 makes FEMALE=Q3 more likely to happen. 
