---
title: "MA 710 - Assignment 2"
author: "MA 710 - Dataminers"
date: "February 22, 2017"
output: html_document
---

### Loading libraries and data
```{r, message=FALSE, warning=FALSE}
library(dplyr)
library(readr)
library(arules)
```

We read in the csv file `MERGED2014_15_PP.csv` into the data frame `csb.df` using the `read_csv` function of the `readr` package.

```{r,message=FALSE, warning=FALSE}

year = 2000
csb.df = data.frame()

for (i in c(1:15)) {
  if (i<10){
    num = paste0("0", i, sep="")
  }
    else {
      num = i
    }
  file_name = paste0('MERGED',year,"_", num, "_PP.csv", sep="")
  csb.temp.df = read_csv(paste0('C:/Users/sevda/Documents/Bentley/710/Assignment 1/CollegeScorecard_Raw_Data/CollegeScorecard_Raw_Data/', file_name))
  csb.temp.df$YEAR = year
  csb.df = rbind(csb.df, csb.temp.df)
  year = year+1
  
}

```

### Data cleaning and variable modification

Based on our decision to select certain variables, we will use the `SELECT` command from `dply` package to create a smaller data frame which contains only the variables of interest.
```{r, message=FALSE, warning=FALSE}
csb.df %>%
  select(CONTROL, ST_FIPS, FAMINC, FEMALE, MARRIED,DEP_STAT_PCT_IND, FIRST_GEN, PREDDEG, CDR3, CDR3_DENOM,PCTPELL, PCTFLOAN, 
         DEBT_N, CUML_DEBT_N, HIGHDEG, REGION, NUMBRANCH, CCBASIC, CURROPER, YEAR) %>%
  {.} -> csb.vars.df
```


To prepare data for association rules, we need to create factor varibles from all of the numeric variables we chose to work with. Because firstly we need to conver the variable type into `factor` and then use the `make.ntiles` function, here we defined a new function `numeric_to_factor` to combine two steps together.
```{r, message=FALSE, warning=FALSE}
make.ntiles = function (inputvar, n) {
  inputvar %>%
    quantile(.,
             (1/n) * 1:(n-1),
             na.rm=TRUE
             ) %>%
    c(-Inf, ., Inf) %>%
    cut(inputvar,
        breaks=.,
        paste("Q", 1:n, sep="")
        )
}

numeric_to_factor <- function(var){
  var.temp= as.numeric(var)
  var.F=make.ntiles(var.temp, 10)
  return(var.F)
}

csb.temp.df %>%
  mutate(FAMINC=numeric_to_factor(FAMINC),
         FEMALE=numeric_to_factor(FEMALE),
         MARRIED=numeric_to_factor(MARRIED),
         DEP_STAT_PCT_IND=numeric_to_factor(DEP_STAT_PCT_IND),
         FIRST_GEN=numeric_to_factor(FIRST_GEN),
         CDR3=numeric_to_factor(CDR3),
         CDR3_DENOM=numeric_to_factor(CDR3_DENOM),
         PCTPELL=numeric_to_factor(PCTPELL),
         PCTFLOAN=numeric_to_factor(PCTFLOAN),
         DEBT_N=numeric_to_factor(DEBT_N),
         CUML_DEBT_N=numeric_to_factor(CUML_DEBT_N),
         NUMBRANCH=as.factor(NUMBRANCH),
         YEAR = as.factor(YEAR),
         CURROPER = as.factor(CURROPER),
         CCBASIC = as.factor(CCBASIC),
         HIGHDEG = as.factor(HIGHDEG)
        ) %>%
  {.} -> csb.fact.df
```
For the last step, we need to make sure that all variables we used here are categorical variables using following code:
```{r}
csb.fact.df %>%
  sapply(as.factor) %>%
  as.data.frame %>%
  {.} -> csb.var.df
```

The data frame `csb.var.df` contains our final set of variables, ready for further analysis.

The next step is to create the association rules. We will use the apriori function from the arules package.

```{r}
rules = apriori(csb.var.df, parameter = list(supp = 0.1, conf = 0.7, minlen = 2, maxlen = 10))
print(paste0("Number of rules:", length(rules)))

```

