---
title: "MA 710 - Assignment 2"
author: "Chengdong Liang"
date: "February 22, 2017"
output: html_document
---

### Loading libraries and data
```{r, message=FALSE, warning=FALSE}
library(dplyr)
library(readr)
```

We read in the csv file `MERGED2014_15_PP.csv` into the data frame `csb.df` using the `read_csv` function of the `readr` package.

```{r,message=FALSE, warning=FALSE}
csb.df = read_csv(paste0('C:/Users/Chengdong Liang/Desktop/MA 710/Assignment 1/CollegeScorecard_Raw_Data/', 'MERGED2014_15_PP.csv'))

```

### Data cleaning and variable modification

Based on our decision to select certain variables, we will use the `SELECT` command from `dply` package to create a smaller data frame which contains only the variables of interest.
```{r, message=FALSE, warning=FALSE}
csb.df %>%
  select(CONTROL, ST_FIPS, FAMINC, FEMALE, MARRIED,DEP_STAT_PCT_IND, FIRST_GEN, PREDDEG, CDR3, CDR3_DENOM,PCTPELL, PCTFLOAN, 
         DEBT_N, CUML_DEBT_N, HIGHDEG, REGION, NPT4_PRIV, NPT4_PUB, NUMBRANCH, CCBASIC, CURROPER) %>%
  {.} -> csb.vars.df
```

For variable `NPT4_PRIV` and `NPT4_PUB`, we combine them together and get a new vaiable `tuition` as a target variable. Firstly we convert the class of these two variables from character to numeric. And then eliminate missing values of them because thoes two variables are mutually exclusive using `replace_na_values` function. As the last step, we use `MUTATE` command from the `dplyr` library and `subset` function to alter variables of the input data frame `csb.vars.df`.

```{r, message=FALSE, warning=FALSE}
replace_na_values <- function (variable){
  for (i in 1:length(variable)){
    if (is.na(variable[i])){
      variable[i] = 0
    }
  }
  return (variable)
  }

csb.vars.df %>% 
  mutate(NPT4_PRIV = as.numeric(NPT4_PRIV),
         NPT4_PUB = as.numeric(NPT4_PUB),
         NPT4 = replace_na_values(NPT4_PRIV) + replace_na_values(NPT4_PUB)) %>%
  subset(select= -c(NPT4_PRIV,NPT4_PUB)) %>%
  {.} -> csb.temp.df
```

To prepare data for association rules, we need to create factor varibles from all of the numeric variables we chose to work with. Because firstly we need to conver the variable type into `factor` and then use the `make.ntiles` function, here we defined a new function `numeric_to_factor` to combine two steps together.
```{r, message=FALSE, warning=FALSE}
make.ntiles = function (inputvar, n) {
  inputvar %>%
    quantile(.,
             (1/n) * 1:(n-1),
             na.rm=TRUE
             ) %>%
    c(-Inf, ., Inf) %>%
    cut(inputvar,
        breaks=.,
        paste("Q", 1:n, sep="")
        )
}

numeric_to_factor <- function(var){
  var.temp= as.numeric(var)
  var.F=make.ntiles(var.temp, 3)
  return(var.F)
}

csb.temp.df %>%
  mutate(FAMINC=numeric_to_factor(FAMINC),
         MARRIED=numeric_to_factor(MARRIED),
         DEP_STAT_PCT_IND=numeric_to_factor(DEP_STAT_PCT_IND),
         FIRST_GEN=numeric_to_factor(FIRST_GEN),
         CDR3=numeric_to_factor(CDR3),
         CDR3_DENOM=numeric_to_factor(CDR3_DENOM),
         PCTPELL=numeric_to_factor(PCTPELL),
         PCTFLOAN=numeric_to_factor(PCTFLOAN),
         DEBT_N=numeric_to_factor(DEBT_N),
         CUML_DEBT_N=numeric_to_factor(CUML_DEBT_N),
         NPT4=numeric_to_factor(NPT4),
         NUMBRANCH=numeric_to_factor(NUMBRANCH)
        ) %>%
  {.} -> csb.fact.df
```
For the last step, we need to make sure that all variables we used here are categorical variables using following code:
```{r}
csb.fact.df %>%
  sapply(as.factor) %>%
  as.data.frame %>%
  {.} -> csb.var.df
```

The data frame `csb.var.df` contains our final set of variables, ready for further analysis.
